---
title: "Modeling heterolasmy in mtDNA"
author: "Soumik Purkayastha"
date: "01/23/2020"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Data processing

* Get 700 randomly chosen individuals with count data for 16569 bases. 
* A typical individual has a 4 * 16569 matrix of count data. 
* Transform count matrix to proportion matrix. 
* **Current definition of heteroplasmy: most prevalent base has frequency $<96\%$**

```{r basic, echo = FALSE}
knitr::opts_chunk$set(autodep = TRUE)
suppressMessages(require(tidyverse))
```

# Searching for heteroplasmy hotspots

* For each of the 16569 bases, get individual level information.
* Get count of how many individuals exhibit heterplasmy at that site. 
* Set cut-off of $>100$ individuals to classify a site as a **heteroplasmy hotspot**. 
* Plot of heteroplasmy hotspots given below. 


```{r chunk1, echo = FALSE, cache=TRUE}
data <- suppressMessages(read_tsv("~/Google Drive - Umich/mtDNA/phase5/other/heteroplasmy/sample.txt"))

f <- function(x){
  x <- x/sum(x)
  ifelse(max(x)<0.96, 1, 0)
}

data$pos <- as.factor(data$pos)
data$nwdid <- as.factor(data$nwdid )

hetero.flag <- data %>% group_by(nwdid, pos) %>% summarise(hetero.flag = f(c(NumA, NumC, NumG, NumT)))

hetero.flag <- spread(hetero.flag, pos, hetero.flag)

hetero.sites <- apply(hetero.flag[,-1], 2, sum)

hetero.sites <- data.frame(Y = hetero.sites, 
                           X = 1:16569)

point.x <- hetero.sites$X[hetero.sites$Y>100]
point.y <- rep(0, sum(hetero.sites$Y>100))

ggplot(hetero.sites) + 
  geom_line(aes(x= X, y = Y)) + 
  theme_bw() +
  theme(legend.position = "none") + 
  geom_hline(yintercept = 100, col = "RED", linetype = 2) + 
  geom_point(data = data.frame(point.x, point.y), 
             aes(x = point.x, y = point.y, col="RED")) + 
  labs(title = "Count of individuals exhibiting heteroplasmy", 
       subtitle = "Dots on X-axis show bases where >100 people exhibit heteroplasmy") + 
  xlab("Base position") + 
  ylab("Number of individuals")



sites <- which(hetero.sites$Y>100)
```

# Distribution of bases at heteroplasmy hotspots

* For any heteroplasmy hotspot, select all 'heteroplasmic individuals'. 
* For each person, calculate proportion of each of the 4 bases. 
* For each base, get density plot of the calculated proportions. 
* Generate overlaid version of the 4 density plots obtained.

An example is given below. 

```{r chunk2, echo = FALSE, cache = FALSE}
prop <- function(x){
  return(x/sum(x))
}

site.id <- sample(sites, 1)

hetero.people <- hetero.flag[(hetero.flag[,(site.id + 1)] == 1),1]

dist.data <- suppressMessages(inner_join(data %>% filter(pos == site.id), hetero.people))

prop.data <- t(apply(dist.data[,c(3:6)], 1, prop))

dist.data <- dist.data %>% add_column(PropA = prop.data[,1], 
                                      PropC = prop.data[,2],
                                      PropG = prop.data[,3],
                                      PropT = prop.data[,4])

dist.data <- pivot_longer(dist.data[,-c(3:6)], 
             cols = c(PropA, PropC, PropG, PropT), 
             names_to = "Base")

dist.data$Base <- ifelse(dist.data$Base == "PropA", "A", 
                         ifelse(dist.data$Base == "PropC", "C",
                                ifelse(dist.data$Base == "PropT", "T","G")))


ggplot(dist.data, 
       aes(x = value, fill = Base)) + 
         geom_density(alpha = 0.4) + 
  theme_bw() + 
  labs(title = paste("Overlaid density plots for each base for heteroplasmy hotspot", site.id))
```





